---
title: "Reconstructing Modern Diets Using Isotopic Hair Analysis and Isotope Ratio Geographical Equivalents"
author: "Anne Poole, Aoife Campbell-Smith, Jesus Figueroa Alcantar, Michelle Henry, Zach Bradford"
date: now
format: 
  html:
    embed-resources: true
execute: 
  warning: false
  self-contained: true
---


## Plotting Individual δ15N (‰) and δ13C (‰) Archy 486 Sample ID's by C3 Terrestrial Ecosystem and Individual Diet Classification


```{r}
# Loading Packages
library(tidyverse)
library(readxl)
library(dplyr)
library(ggrepel)
library(here)
```

The code blocks below produce Figure 1 on our poster:

```{r}
# read sample and global data into variables

sample_data <- 
  read_csv(here("data/sample_CN.csv"))

global_data_indiv <- 
  read_excel(here("data/Global N15 Values.xlsx"), 
             sheet ="Modern subsistence ind.")

global_data_av <- 
  read_excel(here("data/Global N15 Values.xlsx"), 
             sheet ="modern subsistence averages") %>% 
  drop_na(`δ13C_MDE (‰)`)

# read in British and Michigan Data

British_data <- read_csv(here("data/British.csv"),
                         col_names = FALSE)

Michigan_data <- read_csv(here("data/Michigan_dataset .csv"),
                         col_names = FALSE)
  

```

```{r}
#| label: fig-1
#| fig-cap: "Scatterplot of individual δ15N (‰) and δ13C (‰) Archy 486 sample ID’s grouped by C3 Terrestrial Ecosystem and individual diet classification (Müldner, G. (2018). Plot produced by Anne and Aoife. "
  # plot our hair data
library(ggrepel)
ggplot(sample_data) +
    theme_minimal() +
  geom_point(aes(x = d.13C.12C.VPDB ,
                 y = d.15N.14N.air), 
             size = 3,
             colour="red") +
  # label each data point with sample ID
  geom_text_repel(
    aes(x = `d.13C.12C.VPDB`,
        y = `d.15N.14N.air`,
        label = Identifier.1),
    size=3,
    # label all ID's despite overlap
    max.overlaps = Inf) +
  # label x and y axis for consistency with other plots
  labs(x = "δ13C (‰)",
       y = "δ15N (‰)",
       size=3,
       linewidth=2)+
  # create outlines for herbivores, omnivores, carnivores using parameters from Müldner, G. (2018).
  annotate(
    "rect",
    xmin = c(-22, -21, -20),
    xmax = c(-20, -19, -18),
    ymin = c(4, 6, 8),
    ymax = c(5.9, 7.9, 10),
    fill = NA,
    color = "black",
    linewidth=0.5
  ) +
  # create outline for "Terrestrial C3 Ecosystem" for context on graph.
  annotate(
    "rect",
    xmin = c(-23),
    xmax = c(-17),
    ymin = c(1),
    ymax = c(13),
    color = "Dark Green",
    fill = NA,
    linewidth=1
  ) +
  # label all outlines for context and clarity
  annotate(
    "text",
    x = -20,
    y = 13.5,
    label = "",
    color = "Dark Green",
    parse = TRUE,
    size=5
  ) +
  annotate(
    "text",
    x = -21.5,
    y = 6.5,
    label = "Herbivores",
    parse = TRUE,
    size=5
  ) +
  annotate(
    "text",
    x = -20.5,
    y = 8.5,
    label = "Omnivores",
    parse = TRUE,
    size=5
  ) +
  annotate(
    "text",
    x = -19.5,
    y = 10.5,
    label = "Carnivores",
    parse = TRUE,
    size=5
  ) +
  theme_minimal(base_size = 14) +
    xlab(expression(italic(delta)^13*C*" (‰)")) +
  ylab(expression(italic(delta)^15*N*" (‰)")) 
# +
 #scale_x_continuous(sec.axis = dup_axis(
    #name = "% C4")) +
  #labs(title = "Comparing Archy 486 Samples to Trophic Levels in a Terrestrial C3 Ecosystem") 


ggsave(here("figures",
            "trophic_level.png"),
       bg ="white",
       h = 5, # experiment with h and w to get the right size and proportion
       w = 9,
       units = "in",
       dpi = 900) # make the image nice and crisp))

```

## Plotting Average δ15N (‰) and δ13C (‰) for ARCHY 486 Class and Modern Global Subsistence Average δ15N (‰) and δ13C (‰) (with line range for standard deviation)

The code blocks below produce Figure 2 on our poster:

```{r}

global_data <- 
  read_excel(here("data/Global N15 Values.xlsx"),
             sheet = "Modern subsistence ind.")
global_values <- 
  read_excel(here("data/Global N15 Values.xlsx"),
             sheet = "Modern non subsistence averages")

global_subsistence_values <- 
  read_excel(here("data/Global N15 Values.xlsx"),
    sheet = "modern subsistence averages")
```

```{r}
global_subsistence_values_labs <-
  global_subsistence_values %>%
  select(`location`, `δ15N_MDE (‰)`, `SD...6`, `δ13C_MDE (‰)`, `SD...4` ) %>%
  group_by(`location`) %>%
  drop_na()

# rename col
renamed_subsistence <-
global_subsistence_values_labs %>%
  rename(`δ13C (‰)` = `δ13C_MDE (‰)`,
         `δ15N (‰)` = `δ15N_MDE (‰)`) %>%
drop_na()

# select relevant columns from our data
sample_data_perc <-
sample_data %>%
  select(`Identifier.1`, `d.13C.12C.VPDB`, `d.15N.14N.air`) %>% 
  group_by(`Identifier.1`)
  

# select relevant columns from global data
global_values_perc <-
global_values %>%
  select(`location...1`, `δ15N_MDE (‰)`, `δ13C_MDE (‰)`) %>%
  group_by(`location...1`) %>%
  drop_na()

# rename columns from our data
sample_data_perc_lab <-
sample_data_perc %>%
  rename(`δ13C (‰)` = `d.13C.12C.VPDB`,
         `δ15N (‰)` = `d.15N.14N.air`)

 # create a df for the means of `δ13C (‰)` `δ15N (‰)` ratios for class and global data

 `δ13C (‰)` <-
   mean(sample_data_perc_lab$`δ13C (‰)`)
`δ15N (‰)`<-
   mean(sample_data_perc_lab$`δ15N (‰)`)
 
 Statistics_isotopes <- data.frame(means = c("mean_C13", "mean_N15"), Statistics = c(`δ13C (‰)`, `δ15N (‰)`))

mean_sd_C13 <-
  sd(sample_data_perc_lab$`δ13C (‰)`)
mean_sd_N15 <-
  sd(sample_data_perc_lab$`δ15N (‰)`)

# create a class data average dataframe
Stats_isotopes <-
  Statistics_isotopes %>% 
  mutate(mean_sd_C13, mean_sd_N15)

# rename columns in global averages dataframe
global_values_perc_lab <-
global_values_perc %>%
  rename(`δ13C (‰)` = `δ13C_MDE (‰)`,
         `δ15N (‰)` = `δ15N_MDE (‰)`)
  
```

```{r, fig.width = 10}
#| label: fig-standard-13C-15N
#| fig-cap: "Average δ15N (‰) and δ13C (‰) for class hair samples with line range for standard deviation relative to modern global subsistence average δ15N (‰) and δ13C (‰) with line range for standard deviation (Bird et al. 2021). Plot produced by Aoife and Anne."

library(ggrepel)
ggplot() +
    theme_minimal() +
  # our data plot
  geom_point(data = Stats_isotopes,
             aes(x = `δ13C (‰)`,
                 y = `δ15N (‰)`),
             size = 4,
              color = "red") +
  # our error bar
    geom_linerange(data =  Statistics_isotopes,
                (aes(x = `δ13C (‰)`,
                y = `δ15N (‰)` ,
                ymin=`δ15N (‰)` -1*mean_sd_N15, 
                ymax=`δ15N (‰)` +1*mean_sd_N15)),
              color = "red") +
    # our error bar
    geom_linerange(data =  Statistics_isotopes,
                (aes(x = `δ13C (‰)`,
                y = `δ15N (‰)` ,
                xmin=`δ13C (‰)` -1*mean_sd_C13, 
                xmax=`δ13C (‰)` +1*mean_sd_C13)),
              color = "red") +
  # plot global data
  geom_point(data = renamed_subsistence,
             aes(x = `δ13C (‰)`,
                 y = `δ15N (‰)`,
                 color = `location`), 
             size = 3) +
  geom_linerange(data = renamed_subsistence,
                (aes(x = `δ13C (‰)`,
                y = `δ15N (‰)` ,
                ymin=`δ15N (‰)` -1*SD...6, 
                ymax=`δ15N (‰)` +1*SD...6, 
                color = `location`))) +
  geom_linerange(data = renamed_subsistence,
                (aes(x = `δ13C (‰)`,
                y = `δ15N (‰)` ,
                xmin=`δ13C (‰)` -1*SD...4, 
                xmax=`δ13C (‰)` +1*SD...4, 
                color = `location`))) +
  theme_minimal(base_size = 14) +
    xlab(expression(italic(delta)^13*C*" (‰)")) +
  ylab(expression(italic(delta)^15*N*" (‰)")) 

ggsave(here("figures", 
            "geography.png"),
       bg ="white",
       h = 5, # experiment with h and w to get the right size and proportion
       w = 9,
       units = "in",
       dpi = 900) # make the image nice and crisp))

```


## ARCHY 486 Class Diets Survey Analysis
The code blocks below produce Figure 3 on our poster:

```{r}
eating_habits_data <- read_excel(here("data/Food Survey Responses-2.xlsx"))

```

```{r}
eating_habits_data %>%
group_by(sample_id)
```

```{r eval=FALSE}
#| label: fig-food-survey
#| fig-cap: "Eating habits surveyed by Archy 486 students Sample ID numbers. Plot produced by Jesus and Zach."

 # skipping for now



meat_column <- rep(eating_habits_data$meat_q, each = 9)
fish_column <- rep(eating_habits_data$fish_q, each = 9)
dairy_column <- rep(eating_habits_data$dairy_q, each = 9)
soda_column <- rep(eating_habits_data$soda_q, each = 9)
fast_food_column <- rep(eating_habits_data$fastfood_q, each = 9)

sample_id_column <- rep(eating_habits_data$sample_id, each = 9)

data <- data.frame(
  Question = rep(c("Meat", "Fish", "Dairy", "Soda", "Fast_Food"), each = 9),
  Frequency = c(meat_column, fish_column, dairy_column, soda_column, fast_food_column),
  Sample_ID = factor(sample_id_column))

unique_sample_ids <- unique(data$Sample_ID)

color_palette <- colorRampPalette(c("red", "blue", "green", "purple", "orange"))(length(unique_sample_ids))

ggplot(data, aes(x = Sample_ID, y = Frequency, color = Sample_ID)) +
  geom_point(stat = "identity", position = "dodge", size = 8, shape = 18) +
  labs(title = "Eating Habits Comparison in ARCHY 486 Class",
       x = "Sample ID",
       y = "Frequency") +
  scale_fill_manual(values = setNames(color_palette, unique_sample_ids)) +
  facet_wrap(~Question, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(here("analysis",
            "figures", 
            "survey_results.png"),
       bg ="white",
       h = 15, # experiment with h and w to get the right size and proportion
       w = 25,
       units = "in",
       dpi = 900) # make the image nice and crisp))


```



```{r}
# subsistence type

#fixing typos in original global data dataset!!!
global_data_renamed <- global_data %>% 
 rename(`Subsistence Style` = `subsistence style`) 


  global_data_right <- global_data_renamed %>% 
  mutate("Subsistence Style" = case_when(
    `Subsistence Style` == "innuit" ~ "Inuit",
    `Subsistence Style` == "rural mongolian" ~ "Rural Mongolian",
    `Subsistence Style` == "Farmers who fish" ~ "Farmers who fish",
    `Subsistence Style` == "villagers" ~ "Villager",
    `Subsistence Style` == "subsistence" ~ "Subsistence",
    `Subsistence Style` == "Pastoralist" ~ "Pastoralist",
    `Subsistence Style` == "Peasant Farmer" ~ "Peasant Farmer",
    `Subsistence Style` == "hunter gatherer" ~ "Hunter Gatherer",
    `Subsistence Style` == "horticulturalists" ~ "Horticulturalist"
))
  
  ###
  
ggplot() +
    theme_minimal() +
  geom_point(data = sample_data,
             aes(x = `d.13C.12C.VPDB`, 
                 y = `d.15N.14N.air`),
             size = 3,
             colour = "red") +
  geom_point(data = global_data_right,
             aes(x = `δ13C_MDE (‰)`, 
                 y = `δ15N_MDE (‰)`,
                 colour = `Subsistence Style`),
             size=1) +
  geom_text_repel(data = sample_data,
             aes(x = `d.13C.12C.VPDB`, 
                 y = `d.15N.14N.air`,
                 label = Identifier.1),
             # label all ID's despite overlap
             max.overlaps = Inf,
             size = 4,
             bg.color = "white",
             bg.r = 0.1) +
  xlim(-30,-10) +
  xlab(expression(italic(delta)^13*C*" (‰)")) +
  ylab(expression(italic(delta)^15*N*" (‰)"))
  #scale_colour_viridis_d()
  #scale_color_brewer(palette = "Dark1")


  ggsave(here("figures", 
            "subsistence_style.png"),
       bg ="white",
       h = 5, # experiment with h and w to get the right size and proportion
       w = 9,
       units = "in",
       dpi = 900) # make the image nice and crisp))

```

```{r}
# British and Michigan comparison

british_and_american_raw <- 
bind_rows(list( oxford = British_data,
                michigan = Michigan_data),
          .id = "location") %>% 
    rename("δ13C (‰)" = X1, 
         "δ15N (‰)" = X2) %>% 
  bind_rows(

sample_data %>% 
  select(`δ13C (‰)` = `d.13C.12C.VPDB`, 
         `δ15N (‰)` = `d.15N.14N.air`) %>% 
  mutate(location = 'uw')
) %>% 
  mutate(location = factor(location, 
                              levels = c('uw', 'michigan', 'oxford')))
  
# find mean and sds

british_and_american_means_and_sd <- 
  british_and_american_raw %>% 
  group_by(location) %>% 
  summarise(mean_C = mean(`δ13C (‰)`),
            mean_N = mean(`δ15N (‰)`),
            sd_C =   sd(`δ13C (‰)`),
            sd_N =   sd(`δ15N (‰)`))

ggplot() +
  theme_minimal() +
  # individual data points
  geom_point(data = british_and_american_raw,
             aes(x = `δ13C (‰)`, 
                 y = `δ15N (‰)`,
                 colour = location),
             size = 2) +
  # mean data points 
  geom_point(data = british_and_american_means_and_sd,
             aes(x = mean_C, 
                 y = mean_N,
                 colour = location),
             size = 5) +
  # error bars
   geom_linerange(data =  british_and_american_means_and_sd,
                (aes(x = mean_C,
                     y = mean_N ,
                 colour = location,
                ymin= mean_N -1* sd_N, 
                ymax=mean_N  +1* sd_N))) +
     geom_linerange(data =  british_and_american_means_and_sd,
                (aes(x = mean_C,
                     y = mean_N ,
                 colour = location,
                xmin= mean_C -1*  sd_C, 
                xmax= mean_C  +1* sd_C))) +
  scale_color_brewer(palette = "Set1",
                     direction=-1,
                     name = "Student\ngroup") +
  xlab(expression(italic(delta)^13*C*" (‰)")) +
  ylab(expression(italic(delta)^15*N*" (‰)"))
  
  
 ggsave(here("figures", 
            "compared_study.png"),
       bg ="white",
       h = 5, # experiment with h and w to get the right size and proportion
       w = 9,
       units = "in",
       dpi = 900) # make the image nice and crisp))

```

## Access Poster
Our poster can be viewed online here:

https://docs.google.com/presentation/d/1MepuOgSIVt9crcZgkjT2MWBexg-ae8j1/edit#slide=id.p1

## Access Data Files
The data files for this project can be downloaded at:

Food Survey
https://docs.google.com/file/d/1P9KIDbB3Hdk1c9_igDC_5hZ4Y_I3svhh/edit?usp=docslist_api&filetype=msexcel

Global N15 Values
https://www.pnas.org/doi/10.1073/pnas.2024642118

Archy 486 Class Samples Isotope Analysis
https://canvas.uw.edu/courses/1687342

## References
Ammer, Saskia, Eric Bartelink, Jennifer Vollner, Bruce Anderson, and Eugénia Cunha. 2020.

“Socioeconomic and Geographic Implications from Carbon, Nitrogen, and Sulfur Isotope Ratios in Human Hair from Mexico.” Forensic Science International 316. Ireland: Elsevier B.V: 110455–110455. doi:10.1016/j.forsciint.2020.110455.

Bataille, Clement P., Michelle MG Chartrand, Francis Raposo, and Gilles St-Jean. 2020. "Assessing geographic controls of hair isotopic variability in human populations: A case study in Canada." PloS one 15, no. 8: e0237105.

Bird, M. I., Crabtree, S. A., Haig, J., Ulm, S., & Wurster, C. M. 2021. A global carbon and nitrogen isotope perspective on modern and ancient human diet. Proceedings of the National Academy of Sciences, 118(19), e2024642118. https://doi.org/10.1073/pnas.2024642118.

Cotton, Jennifer M., and Nathan D. Sheldon. 2013. “Using Stable Carbon and Nitrogen Isotopes of Hair to Teach About Sustainable Agriculture Through Active Learning.” Journal of Geoscience Education 61 (1). Bellingham: National Association of Geoscience Teachers: 59–67. doi:10.5408/12-309.1. https://www.tandfonline.com/doi/full/10.5408/12-309.1.

Mora, Alice. 2022. “Stable Carbon and Nitrogen Isotope Analysis of Archaeological Human Hair: Reconstructing Diet and Health of Ancient Individuals.” Journal of Archaeological Science: Reports 43 (June): 103439. https://doi.org/10.1016/j.jasrep.2022.103439.

Müldner, G. (2018). Investigating medieval diet and society by stable isotope analysis of human bone. In Reflections: 50 Years of Medieval Archaeology, 1957-2007: No. 30 (pp. 327-346). Routledge.

Roy, Diana M., Roberta Hall, Alan C. Mix, and Robson Bonnichsen. 2005. "Using stable isotope analysis to obtain dietary profiles from old hair: A case study from Plains Indians." American Journal of Physical Anthropology: The Official Publication of the American Association of Physical Anthropologists 128, no. 2: 444-452.

